---
title: "Final Project"
author: "Zifeng(Robin) Zhan"
date: "2025-04-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(survival)
library(survminer)
library(dplyr)
library(purrr)
library(broom)
```

# Inspect Raw Data
```{r}
# Load data set
colon <- survival::colon
## quick structure
str(colon, give.attr = FALSE)

## how much is actually missing?
na_totals <- sapply(colon, \(x) sum(is.na(x)))
na_totals
round(na_totals[na_totals > 0] / nrow(colon) * 100, 2)
# nodes has 1.94% missing value and differ has 2.48% missing values
# missingness is small, delete the obs
```

```{r}
# Delete rows with NA
colon <- colon[complete.cases(colon), ]

# Convert integer to factor
colon_clean <- colon %>%
  mutate(sex = factor(sex, labels = c("Female", "Male")),
         obstruct = as.factor(obstruct),
         perfor = as.factor(perfor),
         adhere = as.factor(adhere),
         differ = factor(differ, 1:3, labels = c("well", "moderate", "poor")),
         node4 = as.factor(node4),
         surg = factor(surg, 0:1, labels = c("short", "long")),
         extent = as.factor(extent),
         etype = as.factor(etype))
```
 
# Death Event
```{r}
# Data set for death event
# Filter the data for death event
colon_death <- colon_clean[colon_clean$etype == 2, ]
## ensure one row for each id
stopifnot(all(!duplicated(colon_death$id)))
```

## KP
```{r}
# Fit the Kaplan-Meier model for the death event
km_fit_death <- survfit(Surv(time, status) ~ 1, data = colon_death)

# Plot the Kaplan-Meier curve for death events
ggsurvplot(km_fit_death, data = colon_death, 
           title = "Kaplan-Meier Survival Curve for Death Events",
           xlab = "Time Until Death (Days)",
           surv.median.line = 'hv',
           break.time.by = 500)

# Median survival time
Death_med <- surv_median(km_fit_death)
print(Death_med)
```

The Kaplan-Meier curve declines slowly and almost linearly over the $3000$ days 
follow-up and the median survival time is $2593$ days. At the beginning of the 
study before one year (365 days), the survival probability is roughly above $90\%$, 
which indicate that patients in the study begin with a near-perfect chance of 
remaining alive. Also, the numerous tick marks in the late tail indicate that 
many individuals were censored alive at the later stage of the study, which is 
common because it is hard to follow-up for a long period. 

## Cox
```{r}
# Coxph model with rx(treatment) as covariate
cox_death <- coxph(Surv(time, status) ~ rx, data = colon_death)
# Create fit for different treatment
fit_death <- survfit(cox_death, newdata = data.frame(rx = c("Obs", "Lev", "Lev+5FU")))

# Plot fit for coxph
ggsurvplot(fit_death, data = colon_death, conf.int = TRUE, 
           ylab = "Survival Probability", 
           xlab = "Time Until Death (Days)", 
           title = "Coxph of Death Event by Treatment",
           legend.title = "Treatment",
           legend.lab = levels(colon_death$rx),
           break.time.by = 500)

# median survival time
cox_med <- surv_median(fit_death)
print(cox_med)

# Summary of PH model
summary(cox_death)

```
This plot show the survival curves for three treatment after fitting a Cox model 
with only treatment (`rx`) as the covariate. `Lev+5FU` (blue) curve locate 
above the other two curves, which might indicate that Levamisole+5-FU (`Lev+5FU`) can increase 
patients' survival rate. And the its survival probability decrease from $1$ and 
end above $0.5$, show that most of patient survive after the study. 

`Lev` (green) and `obs` (red) lines do not show much difference. 
The median survival time for `obs` is $2052$ days and for `Lev` is $2257$ days 
greater than `obs`. $95\%$ confidence interval of median survival time for `obs` 
is ($1550$, $2718$) and the lower bound for confidence interval for `Lev` is 
$1767$. Since the two confidence interval is overlap, there is not statistically 
significant different between the median survival time of `obs` and `Lev`. 

Since the $p$-value of likelihood ratio test is $0.003$ less than 
$\alpha=0.05$, there is sufficient evidence to conclude that `rx`
has significant impact on the survival time of patients. 

Since hazard ratio for `Lev` is $0.9392$, patients on `Lev` has $6.08\%$ lower 
hazard rate than observation. Also the $95\%$ confidence interval for `Lev` is 
($0.7524$, $1.1725$) including $1$. Thus, Levamisole does not have significant impact on 
the survival probability.

The hazard ratio for `Lev+5FU` is $0.6819$, `Lev+5FU` has $32\%$ lower hazard rate 
than `Observation`. Also, $95%$ confidence interval ($0.5379$, $0.8646$) does 
not include $1$. Treatment `Lev+5FU` significantly lower the hazard rate and 
increase the survival probability of patient. 

## AIC
```{r}
# 1st Covariate
# List of Covariate to test
uni_vars <- c("obstruct", "adhere", "nodes", "node4", "differ",
              "extent", "surg", "perfor", "age", "sex")

## 2.  Build one model per variable
uni_models <- map(uni_vars, \(v)
  coxph(as.formula(paste("Surv(time, status) ~ rx + ", v)),
        data = colon_death)
) |> set_names(uni_vars)          

## 3.  Grab AIC 
aic_tbl <- map_dbl(uni_models, AIC) |>
           sort() |>
           round(2)

aic_tbl
# 1st = node4
```

```{r}
# 2nd Covariate
uni_vars2 <- c("obstruct", "adhere", "differ",
              "extent", "surg", "perfor", "age", "sex")

uni_models2 <- map(uni_vars2, \(v)
  coxph(as.formula(paste("Surv(time, status) ~ rx + node4 + ", v)),
        data = colon_death)
) |> set_names(uni_vars2)          

aic_tbl2 <- map_dbl(uni_models2, AIC) |>
           sort() |>
           round(2)

aic_tbl2
# 2nd = extent
```


```{r}
# 3rd Covariate
uni_vars3 <- c("obstruct", "adhere", "differ",
               "surg", "perfor", "age", "sex")

uni_models3 <- map(uni_vars3, \(v)
  coxph(as.formula(paste("Surv(time, status) ~ rx + node4 + extent + ", v)),
        data = colon_death)
) |> set_names(uni_vars3)          

aic_tbl3 <- map_dbl(uni_models3, AIC) |>
           sort() |>
           round(2)

aic_tbl3
# 3rd  = surg
```

```{r}
# 4th Covariate
uni_vars4 <- c("obstruct", "adhere", "differ",
               "perfor", "age", "sex")

uni_models4 <- map(uni_vars4, \(v)
  coxph(as.formula(paste("Surv(time, status) ~ rx + node4 + extent + surg + ", v)),
        data = colon_death)
) |> set_names(uni_vars4)          

aic_tbl4 <- map_dbl(uni_models4, AIC) |>
           sort() |>
           round(2)

aic_tbl4
# 4th = differ
```

```{r}
# 5th Covariate
uni_vars5 <- c("obstruct", "adhere", 
               "perfor", "age", "sex")

uni_models5 <- map(uni_vars5, \(v)
  coxph(as.formula(paste("Surv(time, status) ~ rx + node4 + extent + surg + 
                          differ + ", v)),
        data = colon_death)
) |> set_names(uni_vars5)          

aic_tbl5 <- map_dbl(uni_models5, AIC) |>
           sort() |>
           round(2)

aic_tbl5
```

We selected extra covariates by forward AIC while always keeping treatment 
(`rx`) in the model. Adding `node4`, `extent`, and `surg` each cut AIC by $> 2$ 
points, and `differ` lowered it by another 2.4; `obstruct` reduced AIC 
by $< 2$. Because 2 points is the standard threshold for a meaningful gain, 
we stopped at `rx + node4 + extent + surg + differ`. This captures nearly all 
improvement in fit without adding unnecessary parameters.

# Full Model 
```{r}
full_death <- coxph(Surv(time, status) ~ node4 + extent + 
                      surg + differ + rx, data = colon_death)
summary(full_death)
```

After adjusting for the four strongest prognostic factors—`node4`, `extent`, `surg`, 
and `differ`—the overall likelihood-ratio test is highly significant ($p < 2 × 10^{-16}$), 
confirming that the set of covariates is statistically significant to explain variation 
in survival model. From the summary of the cox proptional model, we can observe 
the following effect of treatment and prognostic covariate:

Treatment effect:  
The combination therapy `Levamisole+5-FU`
has statistically significant survival benefit, reducing the hazard of death by 
approximately $31\%$ with (HR $=0.689$, $95\%$ CI $0.54–0.88$). `Levamisole` alone 
does not show significant benefit because the $95\%$ CI ($0.7638-1.1954$) include 
$1$.

Prognostic covariates:  
`node4`: having more than $4$ positive lymph nodes has hazard ratio of $2.4764$ 
and significantly increase the hazard risk by $147\%$ compared to less than $4$ 
lymph nodes ($95\%$ CI $2.03–3.02$).  
`extent`: Contiguous structures of local spread (`extent` = 4) raises the hazard by $377\%$ 
compared to to submucosa of local spread (`extent`=1) (HR = 4.77, 95 % CI 1.96–15.9).  
`surg`: Long time from surgery to registration (`surg` $=1$) also raise the hazard 
rate by $26.18\%$ compared to shorter time (`surg`$=0$) 
(HR $= 1.26$, $95\%$ CI $1.02–1.55$).




